---
title: "Introduction to Probability"
subtitle: Lecture 3
format:
  revealjs:
    theme: clean.scss
    scrollable: true
    footer: "Lecture 3 - [Back to home](https://madfudolig.quarto.pub/introtobiostats/)"
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    slide-number: true
    menu: true
    code-annotations: hover
    chalkboard: true
    engine: knitr
    echo: true
    code-fold: false
    
  pdf: 
    number-sections: true
    engine: knitr
---


```{r setup}
#| echo: false

library(carData)
library(tidyverse)
```

# Outline

- Views of Probability
- Elementary Properties of Probability
- Calculating Probabilities
- Bayes' Theorem
- Screening Tests

# Views of Probability

## Objective Probability

Probability was thought of by statisticians and mathematicians only as an objective phenomenon derived from objective processes. 

::: callout-note
There are two categories of objective probability:

- classical/a priori probability

- frequentist/a posteriori probability

:::

## Classical Probability

Probabilities for random events related to games of chance can be calculated by the processes of abstract reasoning. Therefore, it is not necessary for these events to happen to compute these probabilities.



:::{.callout-note title="Classical/A Priori Probability"}
Probability is defined as follows: If an event can occur in $N$ mutually exclusive and equally likely ways, and if $m$ of these possess a trait $E$, the probability of the occurrence of $E$ is equal to $m/N$.

$$
P(E)= m/N
$$
:::

::: callout-note
As an example, the probability of getting a three after tossing a **fair** six-sided die can be calculated by assuming that each of the six sides is equally likely to be observed. 

The number of outcomes that include a three is 1 out of a total of 6 mutually exclusive outcomes.

Hence, the probability is 1/6.
:::


## Frequentist Probability

The frequentist approach to probability depends on the repeatability of some process and the ability to count the number of repetitions, as well as the number of times that some event of interest occurs. The probability describes the long-run relative frequency of occurrences from the process.

:::{.callout-note title="Frequentist/A Posteriori Probability"}
Probability is defined as follows: If some process is repeated a large number of times, $n$, and if some resulting event with the characteristic $E$ occurs $m$ times, the relative frequency of occurrence of $E$, $m/n$, will be *approximately equal* to the probability of $E$.

$$
P(E) = m/n
$$
:::

::: callout-important
Note that $m/n$ is only an estimate of the probability, and that $n$ should be large to have a better estimate of $P(E)$.

Example: [Monty-Hall Problem](https://www.rossmanchance.com/applets/2021/montyhall/Monty.html)

:::

## Subjective Probability

This view holds that probability measures the confidence that a particular individual has in the truth of a particular proposition. 

::: callout-note
The most famous example of subjective probability is *Bayesian*, which is a mathematically formal method to combine experimental data and expert information in producing probabilities.
:::

# Properties of Probability 

## Helpful Definitions

:::{.callout-note title="Sample Space"}
A sample space is a list of all possible outcomes that might be observed.
:::

:::{.callout-note title="Event"}
An event is any collection of outcomes in the sample space.   
:::

:::{.callout-note title="Mutually Exclusive"}
If two events $E_i$ and $E_j$ contain no elements in common, these two events are said to be disjoint or mutually exclusive.
:::

## Axioms of Probability

The properties of probability was mathematically formalized by Russian mathematician A.N. Kolgomorov by defining the following axioms:

1. For each event $E_i$, the probability of $E_i$ is non-negative.
2. The sum of the probabilities of the mutually exclusive events is equal to 1.
3. Consider any two mutually exclusive events, $E_i$ and $E_j$. The probability of occurrence of either $E_i$ or $E_j$ is equal to the sum of their individual probabilities.



## Calculating Marginal Probabilities

We can calculate the probability of occurrence for events using the sample point method.

:::{.callout-note title="Sample Point Method"}
Under the assumption that every outcome in the sample space is equally likely to occur, the probability that some event $E_i$ occurs is defined as the number of outcomes corresponding to $E_i$ divided by the total number of outcomes in the sample space.

In frequency distributions, this corresponds to the relative frequency of a specific outcome. 
:::

::: callout-important
Probabilities computed using this method are often referred to as **marginal probability**. Marginal probabilities are unconditional and are specific to an event.
:::


## Example

Out of 1,325 university students, 532 of them reported to consume caffeine through coffee. What is the marginal probability of selecting a university student who consumes coffee?

$$
P(E) = \frac{532}{1325}
$$

The resulting probability is `r round(532/1325,3)`. 


## Example 2

Consider the `AMSSurvey` data in the package `carData`. The data set includes the counts of new PhDs in the mathematical sciences for 2008-09 and 2011-12 categorized by type of institution, gender, and US citizenship status. You can learn more about the data set by typing `?carData::AMSsurvey` after installing the `car` package.

::: {.panel-tabset}

### Question + Data
What is the probability of selecting a participant of the survey at random who is not a US citizen in 2008? Use the `count` variable.

```{r}
library(carData)
AMSsurvey
```

### Answer

The total counts can be calculated using the `sum` function. Or simple addition.

```{r}
sum(AMSsurvey$count)
```
We can also count the total number of non-US citizens in the data set.

```{r}
130+29+79+25+89+50+53+39+122+105+28+12
```

Thus, the probability can be calculated as `r 130+29+79+25+89+50+53+39+122+105+28+12`/ `r sum(AMSsurvey$count)` = `r round((130+29+79+25+89+50+53+39+122+105+28+12)/ sum(AMSsurvey$count),3)`

:::








