---
title: "Problem Set 3 Key"
format: 
  html: 
    engine: knitr
    embed-resources: true
  pdf:
    number-sections: true
---

## Problem 1

The table below shows the frequency table of the number of referrals offered for expectant mothers who experience socioeconomic barriers for needed healthcare.

|                         |               |
|-------------------------|---------------|
| **Number of Referrals** | **Frequency** |
| 0                       | 90            |
| 1                       | 132           |
| 2                       | 76            |
| 3                       | 10            |

- What is the probability that a randomly selected participant received at least one referral? [2 pts.]

::: callout-note
We need to calculate the relative frequencies for each category first to calculate the specific probabilities.
:::

```{r}
library(tidyverse)
df <- data.frame(Referrals=c(0,1,2,3),
                 Frequency = c(90,132,76,10))

mutate(df,RelFrequency = Frequency/sum(Frequency))
```

OR

```{r}
total <- sum(c(90,132,76,10))

df <- data.frame(Referrals=c(0,1,2,3),
                 Frequency = c(90,132,76,10),
                 Rel.Frequency = c(90,132,76,10)/total)

df

```



$$
P(X \geq 1) = P(X=1) + P(X=2) + P(X=3) 
$$
or

$$
P(X \geq 1) = 1-P(X=0)
$$

Hence the probability $P(X\leq 1)$ is:

```{r}
1-90/sum(c(90,132,76,10))

# OR 

1-0.2922
```

- What is the expected value of the number of referrals in the sample? [1pt.]

::: callout-note
The expected number of referrals is the sum of the product of the outcome and their corresponding probabilities.

$$
E(X) = 1*(90/308) + 2*(132/308) + 3*(76/308) + 4*(10/308)
$$
:::

```{r}
0*(90/308) + 1*(132/308) + 2*(76/308) + 3*(10/308)

# OR

sum(df$Referrals*df$Rel.Frequency)
```

