---
title: "Problem Set 7 Key"
format: 
  html: 
    engine: knitr
    embed-resources: true
  pdf:
    number-sections: true
---

```{r setup}
library(tidyverse)
library(summarytools)
library(emmeans)
library(effectsize)
```


## Problem 1
The file aaqol_subset.csv includes data from the 2015 Asian American Quality of Life survey implemented in Austin, TX. We are interested to test whether there is a difference in average age across the different Asian ethnicities represented in the study.

```{r}
aaqol <- read.csv("datasets/aaqol_subset.csv")
glimpse(aaqol)
```

a.	What are the statistical hypotheses for testing whether there is a difference in average age across the different Asian ethnicities represented in the study? [2 pts.]

The null hypothesis is that the average age across the different Asian ethnicities are equal. The alternative hypothesis is that there is at least one Asian ethnic group with a different average age.

b.	What is the value of the test statistic and the corresponding number of degrees of freedom? Hint: We might need more than one type of degrees of freedom. [3pts.]

```{r}
mod_aov<- aov(Age~Ethnicity,data=aaqol)
summary(mod_aov)
```


c.	What is the resulting p-value? [1 pt.]

The p-value is 0.0057.

d.	What is the eta-squared value for the ethnicity effect? [1 pt.]

```{r}
eta_squared(mod_aov)
```

The ethnicity effect has an eta squared value of `r round(eta_squared(mod_aov)$Eta2[1],4)`.

e.	Comment on the sufficiency of evidence based on the alternative hypothesis using a significance level of 0.05. Account for the context of the problem. [1 pt.]

At a significance level of 0.05, there is sufficient evidence that at least one Asian ethnic group has a different average age compared to the others.

f.	Perform a post-hoc test involving pairwise comparisons between the average age of each Asian ethnic group. Apply the Tukey correction. Which pair/s of Asian ethnic groups yielded evidence of pairwise difference in average ages? Use a significance level of 0.05. [2pts.]

```{r}
mod_lm <- lm(Age~Ethnicity,data=aaqol)
em_mod_lm <- emmeans(mod_lm,~Ethnicity)
em_mod_lm
contrast(em_mod_lm,method="pairwise",adjust="tukey")
```

At a significance level of 0.05, Asian Indian (39.6) and Korean (45.4) participants yielded evidence of a pairwise difference in average ages, with an estimated difference of -5.82, test statistic of 3.483, and p-value of 0.0047. Other comparisons did not yield any evidence of a difference.

## Problem 2

The data set `Salaries` in the package `carData` includes the 2008-09 nine-month academic salary for Assistant Professors, Associate Professors and Professors in a college in the U.S. Discipline A consists of “theoretical departments”, while Discipline B consists of “applied departments”.

```{r}
library(carData)
glimpse(Salaries)
```


a.	Test whether there is an interaction between rank (Variable `rank`) and discipline (Variable `discipline`). Use a significance level of 0.05. [2pts.]

```{r}
mod_aov <- aov(salary~rank*discipline,data=Salaries)
summary(mod_aov)
```
At a significance level of 0.05, there is no evidence of an interaction effect (p=0.639)

b.	What is the partial eta-squared value for the interaction effect? [1 pt.]


```{r}
eta_squared(mod_aov)
```
The interaction effect has a partial eta squared value of `r round(eta_squared(mod_aov)$Eta2[3],4)`.

c.	What is the partial eta-squared value for the rank effect? [1 pt.]

The rank effect has a partial eta squared value of `r round(eta_squared(mod_aov)$Eta2[1],4)`.

d.	What is the partial eta-squared value for the discipline effect? [1 pts.]

The discipline effect has a partial eta squared value of `r round(eta_squared(mod_aov)$Eta2[2],4)`.


## Problem 3

The data set socmed.csv includes the PHQ-9 scores of freshman and senior students. The students were also asked about their social media use (socmed, levels = low, high) and gaming habits (levels =yes,no) on their electronic devices.

```{r}
socmed <- read.csv("datasets/socmed.csv")
glimpse(socmed)
```


a.	Test whether there is an interaction between social media use and gaming time on their electronic devices. Use a significance level of 0.05. [2pts.]

```{r}

mod_aov <- aov(y~socmed + gaming + socmed:gaming,data=socmed)
summary(mod_aov)

```

There is sufficient evidence of an interaction effect between social media use and gaming habits at the 0.05 significance level. The test statistic was F(1,28)=6.391, and a p-value of 0.0174.

b.	What is the partial eta-squared value for the interaction effect? [1 pt.]

```{r}
eta_squared(mod_aov)
```

The partial eta-square for the interaction effect is `r round(eta_squared(mod_aov)$Eta2[3],2)`.



c.	Are the partial eta-squared values for `socmed` and `gaming` interpretable? Why or why not? (Hint: Main effects are not interpretable if the null hypothesis for the interaction hypothesis test is rejected). [3 pts.]

The partial eta-squared for `socmed` and `gaming` represent the main effects. Because we have evidence of an interaction effect, these effect sizes are not interpretable.

d.	Provide the estimate of the group means for each social media*gaming level combination. [2pts.]

```{r}

mod_lm <- lm(y~socmed + gaming + socmed:gaming,data=socmed)
em_mod_lm <- emmeans(mod_lm,~socmed|gaming)
em_mod_lm
plot(em_mod_lm)
```

e.	Comment on how the estimated group means support your result in (a). [2pts.]

The difference between the group means of `socmed` is higher in the `Yes` gaming group compared to the `No` gaming group, which supports the alternative hypothesis that there is an interaction between these two factors.

